Voting Rights Restoration and Post-Supervision Turnout

### 2019-11-06 markdown/part2_standalone.Rmd
write

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
header-includes:
    - \usepackage{setspace}\doublespacing
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->
\pagenumbering{gobble}
\tableofcontents
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
# Introduction {-}

@@

In all but two states (Maine and Vermont), felony disenfranchisement laws ensure that American citizens convicted of felony offenses lose the right to vote for at least some period of time. In some states, such as Oregon and Massachusetts, individuals lose that right only for the period in which they are actively incarcerated. In other states, notably Kentucky and Iowa, felony convictions result in lifelong disenfranchisement unless a returned citizen receives an individual pardon from the state's governor [@bcj_laws]. This variation in laws flows directly from language in the Fourteenth Amendment which allows states to revoke individuals' voting rights "for participation in rebellion, or other crime." The definition of "other crime," left so vague in the Constitution, is now generally used by states to disenfranchise citizens for any felony offense. The Supreme Court, in cases such as *Richardson v. Ramirez* (1974), has upheld states' right to do just that. Collectively, these laws disenfranchise as many as `r 6.1 - 1.4` million American citizens. Of these, the majority are no longer incarcerated, but are living and working in their communities [@sentencing_2016].^[The figures reported in @sentencing_2016 have been adjusted to reflect the impact of Amendment 4 in Florida.]

@@

Although felony disenfranchisement is highly disempowering for the individuals impacted by the policy, the political consequences of these laws is not entirely clear. The number of incarcerated individuals is relatively low compared to the number of voters. In New York, for instance, `r comma(readRDS("./temp/number_incarcerated.rds"))` individuals were imprisoned in New York State in early 2019, compared with `r round(readRDS("./temp/active_voters_0319.rds")/1000000, digits = 1)` million actively registered voters. Despite the low share of residents who are directly disenfranchised, there is reason to believe the policy impacts more individuals than just those formally disenfranchised. Previous research has demonstrated that felony disenfranchisement reduces turnout even among Black voters whose rights are not stripped. This research has found, in particular, that eligible Black voters are less likely to cast a ballot in states where felony disenfranchisement policies are harsher, an effect often referred to as *de facto* disenfranchisement.

These spillover effects have been identified using a wide variety of empirical methods. @King2016, for instance, leverages state-level variation in disenfranchisement laws to estimate the impact that felony disenfranchisement has on turnout among Black Americans, finding that "African American disenfranchisement plays a unique role in predicting African American voter turnout" (p. 800). @Ochs2006, @Bowers2009, and @Walker2014 similarly exploit state-level differences to estimate the (de)mobilizing effect of felony disenfranchise on eligible voters. As @Bowers2009 sums up: "[I]t is not solely the direct vote of ex-felons that is denied through these laws. [Felony disenfranchisement] impacts the political power of communities that extends beyond felons' collateral penalty" (p. 724). @Burch2013 used neighborhood-level data to further investigate how felony disenfranchisement operates the sub-state level. She finds that "at high concentrations, imprisonment and community supervision have an unequivocally demobilizing effect of neighborhoods" (p. 185).

It is clear that although felony disenfranchisement directly impacts relatively few potential voters, its demobilizing character reaches many citizens who are not formally disenfranchised. Considering the characteristics of the neighborhoods that incarcerated individuals call home --- neighborhoods that are among the most marginalized --- felony disenfranchisement is a large problem.

In this paper, I examine the effect of Executive Order 181 in New York State. Signed in April of 2018, the Executive Order restored voting rights to many individuals on parole. Prior to Executive Order 181, individuals were disenfranchised while incarcerated or on parole. Most individuals who were on parole on Election Day in 2018 were allowed to vote, a major step forward for democracy in the Empire State. In this paper, however, I focus on individuals who were discharged from parole prior to Election Day: individuals who would not have been directly disenfranchised by felony disenfranchisement policies even in the absense of the Executive Order. I hypothesize that Executive Order 181 increased turnout among this population, even though their eligibility was not directly impacted by the policy change.

@@

In the aftermath of the 2000 presidential election, academic interest in the political implications of felony disenfranchisement was stirred thanks to a paper from @Uggen2002. George W. Bush's margin of victory in Florida in 2000 was famously just 537 votes. In their 2002 paper, Uggen and Manza estimate the likely partisan composition of the disenfranchised population with felony convictions in their past. They estimate that, if this group had been allowed to vote, they would have supported Al Gore by a wide margin. Their enfranchisement, Uggen and Manza argued, would have tipped the presidential contest and resulted in the election of Al Gore. They based their estimates on the voting patterns of eligible individuals who were demographically similar to the disenfranchised population. Uggen and Manza demonstrated that felony disenfranchisement can have material political consequences.

In subsequent years, new research called into question Uggen and Manza's assumption that disenfranchised individuals would vote at rates comparable to their sociodemographic peers. In a series of papers between 2009 and 2011, researchers developed methods for directly estimating the turnout of formerly disenfranchised individuals. @Haselswerdt2009 matched release data and voter registration data from Erie County, NY, to estimate turnout among a small group of formerly incarcerated individuals. Traci Burch [-@Burch2010; -@Burch2011] expanded upon this matching methodology to estimate the voting patterns of formerly disenfranchised individuals in a range of states. She used release data from states' Departments of Corrections and their registered voter files to identify formerly incarcerated individuals who went on to register to vote. Using the registered voter files, she was also able to estimate the party affiliation of formerly incarcerated individuals (in states with party registration) and their turnout rates. Her methodology has been used to investigate other questions surrounding the voting patterns of formerly incarcerated individuals under different circumstances and to examine the impact of changes in disenfranchisement policy [@Meredith2013; @Meredith2015].

The causal effect of incarceration on participation is the subject of some debate within the field. Individuals who go to prison share many characterstics with lower propensity voters generally. Less educated citizens, for instance, turnout at low rates whether they have been to prison or not. In an attempt to disentangle sociodemographic characteristics from the experience of imprisonment, @Gerber2017 uses administrative data from Pennsylvania to estimate turnout rates prior to and after incarceration. They argue that the vast majority of low turnout among formerly incarcerated individuals can be explained by observable characteristics, concluding that "it appears that spending time in prison does not have large negative effects on subsequent participation" (p. 1144).

@White2019, however, indicates that interaction with the criminal justice system for individuals in the context of arrests for misdemeanor charges may have depressive effects on turnout. This finding does not necessarily conflict with @Gerber2017; as the earlier paper explains, incarceration often occurs after many other interactions with the criminal justice system. Individuals arrested for misdemeanors, on the other hand, likely reflect a much broader swath of the population --- and, therefore, individuals who may have had fewer interactions with the criminal justice system. The findings in @White2019 agree with much previous research which shows that individuals who have negative interactions with the state are less likely to participate in civic life [@Pierson1993]. @Weaver2010 argues that "contact with the institutions of criminal justice is important in structuring patterns of participation long assumed in the dominant literature to stem primarily from aspects of the individual" (p. 829).

Some research has been done in this area. @Meredith2015 examines the impact of ending permanent disenfranchisement in Iowa. They find that individuals who received letters explicitly informing them of their re-enfranchisement were more likely to cast ballots in the next election than those who did not. @Meredith2013, however, examines states where so-called notification laws went into effect. Although rules about eligibility did not change in these states, new policies required Departments of Corrections to notify formerly disenfranchised individuals of their re-instated voting rights. @Meredith2013 finds no effect from notification in the absense of eligibility changes.

@Gerber2014 conducted a field experiment in Connecticut in advance of the 2012 presidential election, finding that sending mailers to individuals to remind them of their voting rights was successful at increasing turnout among this population. "Whatever the participatory consequences of incarceration," they conclude, "they are not in large part impossible to overcome" (p. 924). It is not clear whether this increase in turnout is *undoing* the depressive effect of incarceration or boosting the participation of individuals whose (low) propensity to vote was unaffected by incarceration.

Executive Order 181 is expected to have increased turnout among formerly incarcerated individuals through two primary mechanisms. The first addresses the impact of negative experiences with the state. To the extent that parole officers are accurately informing their parolees of their newly restored voting rights, Executive Order 181 is likely to bring about a positive interaction between the parolee and the government. Rather than simply have one's rights restored upon completion of sentence, Executive Order 181 may lead parolees to feel explicitly invited back into the democratic process --- an invitation that may be successful at repairing some of the negative associations created through incarceration.

Secondly, Executive Order 181 is expected to dispel confusion about eligibilty. Since @Downs1957, political scientists have argued that an individual's propensity to vote is decided at least in part by comparing the expected costs of casting a ballot with the expected benefits of doing so. The costs of voting while disenfranchised (intentionally or otherwise) are exceedingly high, as high-profile cases in states such as Texas have made clear in recent years [@Flynn2018]. If an individual assessment of the cost of voting is determined in part by the product of their uncertainty about their eligibility and the cost of casting a ballot while ineligible, formerly disenfranchised individals are unlikely to participate unless they are fully certain of their eligibility. As @Drucker2005 and others have detailed, however, many formerly incarcerated individuals are misinformed about their eligibility to cast a ballot. By dispelling uncertainty around eligibility, I hypothesize that Executive Order 181 increased turnout among the formerly disenfranchised.

# Research Design {-}

On April 18^th^, 2018, Governor Andrew Cuomo signed Executive Order 181 which effectively ended the disenfranchisement of New Yorkers on parole. While the Executive Order is obviously beneficial for individuals who are on parole on Election Day (and therefore would have been disenfranchised but for the policy change), the Executive Order may have benefits even for the individuals whose eligibility was not directly impacted. Under the Executive Order, parole officers were required to inform their parolees of their new status under the law. In the analysis that follows, I look only at individuals who were discharged before October 10^th^, 2018 (the registration deadline in New York State). Limiting the analysis to individuals whose eligibility to cast a ballot was not directly impacted by the policy change allows me to explore whether having one's rights restored while still under formal supervision is effective at raising that individual's propensity to vote even once supervision is ended.

In order to estimate the impact of Executive Order 181 on turnout, I begin by exploring whether the turnout rate among individuals discharged from parole after the policy went into effect systematically differed from those discharged earlier. This analysis uses an individual-level logistic regression that estimates individuals' turnout rates and controls for various individual-level characteristics. Such an analysis sheds light on the impact of rights restoration on the turnout rate of the formerly disenfranchised population as a whole.

Such an approach has some benefits: supposing that policymakers care about the political representation of formerly incarcerated residents as a group, then examining the group turnout rate can yield important insights. This approach does not, however, address the impact having one's rights restored has on individual propensity to vote. To interrogate whether an individual's propensity to vote is increased if she has her rights restored prior to discharge from parole, I test whether individuals who had their rights restored prior to discharge from parole turned out at higher rates in midterm election. Because the restoration of voting rights is not randomly assigned, I employ an instrumental variables approach, as discussed in greater detail below.

@@

The criminal justice dataset comes from a public records request filed by the author to obtain individual-level incarceration and parole records for individuals sentenced to incarceration in New York State since 1990. The data includes a host of information, including: first, middle, and last name; date of birth; class of offense; incarceration start and end dates; dates of parole; county of committment; and others. This analysis is limited to individuals incarcerated for felony offenses. Individuals convicted of misdemeanors are not disenfranchised in New York State. These data come from the New York State Department of Corrections and Community Supervision (NYSDOCCS). These data are used to determine when individuals were incarcerated or on parole, and when they finished their parole supervision.

The state does not make a unified database of parolees whose voting rights have been restored available to the public. However, the NYSDOCCS Parolee Lookup website includes a flag indicating whether someone's voting rights have been restored. By using identification numbers from the parolee data obtained from the state, I constructed a list of the individuals who were on parole and had their rights restored.^[Not all parolees listed in the public records request data are included in the lookup tool. For individuals who finished parole between January 1^st^, 2018, and April 17^th^, 2018, 1.0 percent are not in the lookup tool. For those discharged from parole between April 18^th^, 2018, and January 13^th^, 2019 (the latest date of the parole records), 1.2 percent of individuals are not found in the lookup tool.]

@@

Below, I employ a logistic regression to investigate the impact of Executive Order 181 on turnout among all parolees discharged after the policy went into effect. Because not all individuals had their rights restored, I follow this analysis with an instrumental-variables approach to determine the effect of rights restoration on individuals whose rights were in fact formally restored prior to discharge from parole.

## Trends in Turnout {-}
Before analyzing turnout in the 2018 midterms, I begin by examining turnout in the 2016 election. It is possible that individuals discharged from parole shortly before a federal election are more likely to cast a ballot than individuals discharged earlier, whether or not their voting rights were restored. However, as Figure \ref{fig:to-16} makes clear, individuals discharged from parole in the final months before the 2016 presidential election were not substantially more likely to cast a ballot in the election than individuals discharged earlier. The longer an individual has been off of parole, the more likely he is to cast a ballot. For instance, of the individuals last discharged from parole in 2010, `r percent(readRDS("./temp/turnout_in_16_finished_10.rds"), accuracy = 0.1)` cast a ballot in the 2016 election, while just `r percent(readRDS("./temp/turnout_in_16_finished_15.rds"), accuracy = 0.1)` of those last discharged from parole in 2015 did so.^[Figure \ref{fig:to-16} plots individuals' turnout by the last date of discharge from parole. Therefore, individuals discharged from parole in 2010 who reoffended and were discharged from parole again in 2015 are included only in 2015.] Figure \ref{fig:to-16} plots turnout rates by month of parole discharge. A quadratic curve is fitted (weighted by the number of individuals discharged each month), along with a 95 percent confidence band. This curve is fit on monthly data running from January, 2010 through April, 2016, and extended through October, 2016.

```{r to-16-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-16}Turnout in 2016 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to16_chart.rds")
print(j)
```
Figure \ref{fig:to-18} plots month of parole discharge and turnout in the 2018 midterm elections. Once again, a weighted quadratic curve is fitted with a 95 percent confidence band. This curve is fit on monthly data running from January, 2012 through April, 2018, and extended through October, 2018.
```{r to-18-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18}Turnout in 2018 Midterm Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart.rds")
print(j)
```
Figure \ref{fig:to-16} does not indicate that individuals who were discharged from parole shortly before the 2016 presidential election were more likely to cast a ballot than individuals discharged earlier in the year. Figure \ref{fig:to-18}, on the other hand, indicates that New Yorkers discharged from parole in the months leading up to the 2018 election --- many of whom had their rights restored while they were still on parole --- were more likely to participate than those discharged earlier in the year. However, Figures \ref{fig:to-16} and \ref{fig:to-18} are noisy and do not prove that Executive Order increased turnout.

@@

When considered over a multi-year period, the enactment of Executive Order 181 cannot be understood as a natural experiment. The longer an individual has been off of parole, the more likely she is to cast a ballot, but only individuals recently discharged from parole were eligible to have their voting rights restored prior to discharge. For a true natural experiment to hold, an individual's probability of being "assigned" to treatment (here, discharged from parole after the Executive Order went into effect) must be uncorrelated with the outcome of interest (propensity to vote). Figures \ref{fig:to-16} and \ref{fig:to-18} indicate that this is not the case when considering individuals discharged from parole over multiple years.

However, the relationship between time-off-parole and propensity to vote is far weaker in the short term. Figure \ref{fig:to-18-recent} indicates that parole discharge date and turnout rates in the 2018 midterm election are not correlated when we limit the analysis to individuals discharged in 2017 or 2018.

```{r to-18-recent-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18-recent}Turnout in 2018 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart_recent.rds")
print(j)
```
Formalizing this chart into an individual-level logit model demonstrates that time since discharge is not correlated with turnout in the short run. The models in Table \ref{tab:to-18-logit-short} include individuals last discharged from parole between January 1^st^, 2017, and April 17^th^, 2018.

```{r logit-reg-chunk-short, echo = F, results = "asis", message = F}
load("./temp/short_term_to.rdata")

stargazer(date_to_model_notime, date_to_model_time,
          header = F,
          type = "latex",
          covariate.labels = c("Days Since Discharged",
                               "Days Since Discharged\\textsuperscript{2}", "D(Male)", "Age (Years)",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "(\\#tab:to-18-logit-short) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "X", "X"),
                         c("Felony Class Fixed Effects", "X", "X")),
          table.layout = "-dm#-t-a-s-n")
```

The inclusion of time controls in Table \ref{tab:to-18-logit-short} increases the AIC. A Chi-squared test confirms that the model is not improved when controls for time are included.^[[Appendix A] provides further corroboration that being discharged from parole in the months before an election is uncorrelated with propensity to vote by exploring turnout rates in the 2016 presidential election.] When we look only at individuals recently discharged from parole, the length of time an individual has been off parole is not associated with his propensity to vote.

>>>>

---
title: "Voting Rights Restoration and Post-Supervision Turnout"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    # includes: 
    #   in_header: "../code/misc/preamble.tex"
    keep_tex: true
bibliography: "citations.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{rotating}

---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->
\pagenumbering{gobble}
\tableofcontents
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(data.table)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
# Introduction {-}

@@

In all but two states (Maine and Vermont), felony disenfranchisement laws mean that American citizens convicted of felony offenses lose the right to vote for at least some period of time. In some states, such as Oregon and Massachusetts, individuals lose that right only for the period in which they are actively incarcerated. In other states, notably Kentucky and Iowa, felony convictions result in lifelong disenfranchisement unless a returned citizen receives an individual pardon from the state's governor [@bcj_laws]. This variation in laws flows directly from language in the Fourteenth Amendment which allows states to revoke individuals' voting rights "for participation in rebellion, or other crime." The definition of "other crime," left so vague in the Constitution, is now generally used by states to disenfranchise citizens for any felony offense. The Supreme Court, in cases such as *Richardson v. Ramirez* (1974), has upheld states' right to do just that. Collectively, these laws disenfranchise as many as `r 6.1 - 1.4` million American citizens. Of these, the majority are no longer incarcerated, but are living and working in their communities [@sentencing_2016].^[The figures reported in @sentencing_2016 have been adjusted to reflect the impact of Amendment 4 in Florida.]

@@

On April 18^th^, 2018, Governor Andrew Cuomo signed Executive Order 181 which effectively ended the disenfranchisement of New Yorkers on parole. Such a move was of course good for the communities in which parolees live; as discussed above, the disenfranchisement of voters has large spillover effects in the neighborhoods in which these lost voters are concentrated. The change in policy is also beneficial for felony probationers: despite the fact that probationers do not formally lose their voting rights, there is evidence that confusion around the law contributes to *de facto* disenfranchisement among probationers [@Drucker2005]. The executive order is a promising step: by changing the policy to allow all New York citizens living in their communities to cast a ballot, the move has the potential to both re-enfranchise the nearly 30,000 New Yorkers on parole living in the community and to clarify the rules about who is eligible to vote.

Of course, re-extending the right to vote is only the first step. As previous literature has established, interactions with the criminal justice system leaves residents less likely to vote in the future [@White2019]. As @Burch2011 and others have shown, moreover, turnout rates among the formerly incarcerated are extremely low. Formal disenfranchisement policy, the literature has made clear, is just one piece of an interlocking system that serves to disenfranchise minority and marginalized voters [see, for instance, @Fraga2018]. To address *only* the formal laws contributing to disenfranchisement without also interrogating the efficacy of any policy change to boost participation risks leaving much of the system of effective disenfranchisement undisturbed.

There is reason to believe that the executive order may have increased the political participation of formerly disenfranchised individuals. Prior to the policy change, formerly incarcerated individuals had their voting rights restored automatically upon the completion of their parole term. New York's correction code provides no more guidance other than that "upon a person's discharge from community supervision, the department shall notify such person of his or her right to vote and provide such person with a form of application for voter registration" (Section 75 of the New York State Correction Law). Shortly after the implementation of the executive order, Acting Deputy Commissioner for Community Supervision Ana Enright sent a memorandum to New York State parole officers detailing the Department of Corrections' new approach.^[Link: https://nyassembly.gov/member_files/139/webdocs/82103.pdf] The memorandum directs all parole officers to present parolees with voter registration forms and to explain their purpose. Parole officers were also instructed to offer any assistance needed, including help filling out the registration form. In addition to these directives, the memorandum communicates that voter registration is to receive "high priority attention," and it separately calls the program "a **priority** initiative" [emphasis in the original]. Thus, the executive order demands not only that re-enfranchised individuals receive in-person notification of their voting rights, but also that parole officers prioritize their registration.

In the analysis below, I examine the effect of Executive Order 181 on individuals who finished parole before October 10^th^, 2018 (the registration deadline for 2018). These individuals, because they were discharged from parole before the registration deadline, would have been eligible to vote even in the absense of the executive order. This is, of course, only a subset of the individuals who were impacted by Executive Order 181 --- many of the individuals who had their voting rights restored were still on parole on Election Day, and therefore were eligible to vote only because of the rules change. However, by focusing on individuals who would have been eligible to vote either way, I can test whether rights restoration prior to parole discharge serves as an effective encouragement of post-supervision participation.

@@

In a series of papers between 2009 and 2011, researchers developed methods for directly estimating the turnout of formerly disenfranchised individuals. @Haselswerdt2009 matched administrative release data and voter registration data from Erie County, NY, to estimate turnout among a small group of formerly incarcerated individuals. Traci Burch [-@Burch2010; -@Burch2011] expanded upon this matching methodology to estimate the voting patterns of formerly disenfranchised individuals in a range of states. She used release data from states' Departments of Corrections and their registered voter files to identify formerly incarcerated individuals who went on to register to vote. Using the registered voter files, she was also able to estimate the party affiliation of formerly incarcerated individuals (in states with party registration) and their turnout rates. Her methodology has been used to investigate other questions surrounding the voting patterns of formerly incarcerated individuals under different circumstances and to examine the impact of changes in disenfranchisement policy [e.g. @Meredith2013; @Meredith2015].

Much of the literature discussed above has established that formerly incarcerated individuals rarely vote, even when they are no longer formally barred from doing so [@Haselswerdt2009; @Burch2011; @Meredith2015]. The causal effect of incarceration on participation is the subject of some debate within the field. Individuals who go to prison share many characteristics with lower propensity voters generally. Less educated citizens, for instance, turnout at low rates whether they have been to prison or not. In an attempt to disentangle sociodemographic characteristics from the experience of imprisonment, @Gerber2017 uses administrative data from Pennsylvania to estimate turnout rates prior to and after incarceration. They argue that the majority of the low turnout observed among formerly incarcerated individuals can be explained by observable characteristics, concluding that "it appears that spending time in prison does not have large negative effects on subsequent participation" (1144).

@White2019, however, comes to a different conclusion. Using administrative court and voter file records from Harris County, Texas, she finds that individuals assigned jail time for misdemeanor offenses are less likely to participate in future elections than similarly-situated misdemeanants who do not go to jail. This finding does not necessarily conflict with @Gerber2017; as the earlier paper explains, prison often occurs after many other interactions with the criminal justice system. By the time an individual is incarcerated, interactions with the criminal justice system may have already reduced his propensity to vote. Individuals arrested for misdemeanors, on the other hand, likely reflect a much broader swath of the population --- and, therefore, individuals who may have had fewer interactions with the criminal justice system.

Regardless of the precise mechanism, the low turnout among formerly incarcerated individuals is cause for concern, particularly given the racialized aspects of the criminal justice system. A criminal justice system that inflicts serious consequences on a population that has relatively little political voice is problematic. Whether or not incarceration *causes* low turnout the state has a unique opportunity to craft policies that will impact individuals under formal supervision. Even if incarceration does not lead to lower turnout, policies targeting individuals caught up in the criminal justice system might still be effective at increasing turnout for this population.

The literature on turnout among formerly incarcerated individuals leads me to hypothesize that rights restoration prior to parole discharge will boost turnout among formerly incarcerated individuals. I expect that this increased propensity to vote will arise due both thanks to better clarity about individual eligibility to participate, but also because rights restoration may repair some of the distrust of the state engendered by incarceration.

Formerly convicted individuals are very often confused about their eligibility to vote [@Drucker2005; @locked_out]. Some research indicates that dispelling misinformation boosts turnout among the formerly disenfranchised. @Meredith2015 examines the impact of ending permanent disenfranchisement in Iowa. They find that individuals who received letters explicitly informing them of their re-enfranchisement were more likely to cast ballots in the next election than those who did not. @Meredith2013, however, examines states where so-called notification laws went into effect. Although rules about eligibility did not change in these states, new policies required Departments of Corrections to notify formerly disenfranchised individuals of their re-instated voting rights. @Meredith2013 finds no effect on turnout from notification in the absence of eligibility changes. @Gerber2014 conducted a field experiment in Connecticut in advance of the 2012 presidential election, finding that sending mailers to individuals to remind them of their voting rights was successful at increasing turnout among this population. "Whatever the participatory consequences of incarceration," they conclude, "they are not in large part impossible to overcome" (924). Even if incarceration does not decrease individuals' propensity to vote, there is reason to believe that reminding formerly incarcerated individuals of their rights increases participation.

Research also indicates that individuals who have negative interactions with the state are less likely to participate in civic life [@Pierson1993]. @Weaver2010 argues that "contact with the institutions of criminal justice is important in structuring patterns of participation long assumed in the dominant literature to stem primarily from aspects of the individual" (829). To the extent that parole officers are accurately informing their parolees of their newly restored voting rights, Executive Order 181 is likely to bring about a positive interaction between the parolee and the government. Rather than simply have one's rights restored upon completion of sentence, Executive Order 181 may lead parolees to feel individually invited back into the democratic process --- an invitation that may be successful at repairing some of the negative associations created through incarceration. Such repairs may lead to increased propensity to participate in elections in the future.

@@

The criminal justice dataset comes from a public records request filed by the author to obtain individual-level incarceration and parole records for individuals sentenced to incarceration in New York State since 1990. The data includes a host of information, including: first, middle, and last name; date of birth; class of offense; incarceration start and end dates; dates of parole; county of committment; and others. This analysis is limited to individuals incarcerated for felony offenses. Individuals convicted of misdemeanors are not disenfranchised in New York State. These data come from the New York State Department of Corrections and Community Supervision (NYSDOCCS). These data are used to determine when individuals were incarcerated or on parole, when they finished their parole supervision, and demographic information such as age and race.

Following Executive Order 181, the Department of Corrections and Community Supervision began indicating on their online Parolee Lookup Tool whether a parolee had her voting rights restored. By using the identification number provided from the parolee public records request and this website, I was able to identify individuals who had their voting rights restored.^[Not all parolees listed in the public records request data are included in the lookup tool. For individuals who finished parole between January 1^st^, 2018, and April 17^th^, 2018, 1.0 percent are not in the lookup tool. For those discharged from parole between April 18^th^, 2018, and January 13^th^, 2019 (the latest date of the parole records), 1.2 percent of individuals are not found in the lookup tool.] There were `r comma(readRDS("./temp/how_many_restored.rds"))` individuals who were discharged from parole before the registration deadline whose rights were restored while still under supervision.

@@

Before analyzing turnout in the 2018 midterms, I begin by examining turnout in the 2016 election. It is possible that individuals discharged from parole shortly before a federal election are more likely to cast a ballot than individuals discharged earlier, whether or not their voting rights were restored. However, as Figure \ref{fig:to-16} (which plots 2016 turnout rates by month of parole discharge) makes clear, individuals discharged from parole in the final months before the 2016 presidential election were not substantially more likely to cast a ballot in the election than individuals discharged earlier. The longer an individual has been off of parole, the more likely he is to cast a ballot. For instance, of the individuals last discharged from parole in 2010, `r round(readRDS("./temp/turnout_in_16_finished_10.rds") * 100, digits = 1)` percent cast a ballot in the 2016 election, while just `r round(readRDS("./temp/turnout_in_16_finished_15.rds") * 100, digits = 1)` percent of those last discharged from parole in 2015 did so.^[Figure \ref{fig:to-16} plots individuals' turnout by the last date of discharge from parole. Therefore, individuals discharged from parole in 2010 who reoffended and were discharged from parole again in 2015 are included only in 2015.] A quadratic curve is fitted (weighted by the number of individuals discharged each month), along with a 95 percent confidence band. This curve is fit on monthly data running from January 2010 through April 2016, and extended through October 2016.

```{r to-16-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-16}Turnout in 2016 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to16_chart.rds")
print(j)
```
Figure \ref{fig:to-18} plots month of parole discharge and turnout in the 2018 midterm elections. Once again, a weighted quadratic curve is fitted with a 95 percent confidence band. This curve is fit on monthly data running from January 2012 through April 2018, and extended through October 2018.
```{r to-18-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18}Turnout in 2018 Midterm Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart.rds")
print(j)
```
Figure \ref{fig:to-16} does not indicate that individuals who were discharged from parole shortly before the 2016 presidential election were more likely to cast a ballot than individuals discharged earlier in the year. Figure \ref{fig:to-18}, on the other hand, indicates that New Yorkers discharged from parole in the months leading up to the 2018 election --- many of whom had their rights restored while they were still on parole --- *were* more likely to participate than those discharged earlier in the year. However, Figures \ref{fig:to-16} and \ref{fig:to-18} are noisy and do not prove that the executive order increased turnout.

@@

When considered over a multi-year period, the enactment of Executive Order 181 cannot be understood as a natural experiment. The longer an individual has been off of parole, the more likely she is to cast a ballot, but only individuals recently discharged from parole were eligible to have their voting rights restored prior to discharge. For a true natural experiment to hold, an individual's probability of being "assigned" to treatment (here, discharged from parole after the executive order went into effect) must be uncorrelated with the outcome of interest (propensity to vote). Figures \ref{fig:to-16} and \ref{fig:to-18} indicate that this is not the case when considering individuals discharged from parole over multiple years.

However, the relationship between time-off-parole and propensity to vote is far weaker in the short term. Figure \ref{fig:to-18-recent} indicates that parole discharge date and turnout rates in the 2018 midterm election are not correlated forindividuals discharged in 2017 or 2018 before the executive order was enacted.

```{r to-18-recent-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18-recent}Turnout in 2018 Midterm Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart_recent.rds")
print(j)
```
Formalizing this chart into an individual-level logistic model demonstrates that time since discharge is not correlated with turnout in the short run. The models in Table \ref{tab:to-18-logit-short} include individuals last discharged from parole between January 1^st^, 2017, and April 17^th^, 2018 --- that is to say, all individuals discharged prior to the implementation of the executive order.

```{r logit-reg-chunk-short, echo = F, results = "hide", message = F, warning = F}
load("./temp/short_term_to.rdata")

stargazer(date_to_model_notime, date_to_model_time,
          header = F,
          type = "latex", notes.align = "l",
          covariate.labels = c("Days Since Discharged", "Days Since Discharged\\textsuperscript{2}",
                               "D(Male)", "Age (Years)",
                               "Years on Parole"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "\\label{tab:to-18-logit-short} Individual-Level Logit Models",
          table.placement = "H",
          omit = c("race", "felony"),
          add.lines=list(c("Race / Ethnicity FE", "X", "X"),
                         c("Felony Class FE", "X", "X")),
          table.layout = "-dm#-t-a-s-n",
          out = "./temp/test.tex",
          notes = "TO REPLACE",
          out.header = F,
          se = list(
            sqrt(diag(vcovHC(date_to_model_notime, type = "HC0"))),
            sqrt(diag(vcovHC(date_to_model_time, type = "HC0"))))
          )

j <- fread("./temp/test.tex", header = F, sep = "+")

note.latex <- "\\multicolumn{3}{l}{\\scriptsize{\\parbox{.5\\linewidth}{\\vspace{2pt}$^{***}p<0.01$, $^{**}p<0.05$, $^*p<0.1$. \\\\Robust standard errors in parentheses.}}}"

j <- j %>% 
  mutate(V1 = ifelse(grepl("TO REPLACE", V1), note.latex, V1),
         V1 = ifelse(grepl("\\\\#tab", V1), gsub("\\\\#", "", V1), V1)) %>% 
  filter(!grepl("Note:", V1)) %>% 
  mutate(n = row_number())

insert1 <- "\\fontsize{10}{12}\\selectfont"

j <- bind_rows(j, data.frame(V1 = c(insert1), n = c(1.1))) %>% 
  arrange(n) %>% 
  select(-n)

write.table(j, "./temp/table3.tex", quote = F, col.names = F,
            row.names = F)

```
\input{"../temp/table3.tex"}
The inclusion of time controls in Model 2 in Table \ref{tab:to-18-logit-short} increases the AIC. A Chi-squared test confirms that the model is not improved when controls for time are included. When we look only at individuals recently discharged from parole, the length of time an individual has been off parole is not associated with his propensity to vote.^[[Appendix A] provides further corroboration that being discharged from parole in the months before an election is uncorrelated with propensity to vote by exploring turnout rates in the 2016 presidential election.] This indicates that using individuals released in 2017 and 2018 prior to the executive order as a control group is feasible.
